### 2019-01-20 Datelife_paper1.Rmd
some figures

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
figure_runtime_main <- "Fig. 1"  # make_datelife_query_plus_get_datelife_result
```

@@

<!--INTRODUCTION-->
Time of lineage divergence<!--named biological entities, taxon ages?--> constitutes in many ways the fundamental/main knowledge necessary for evolutionary understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
Along with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
Coupled to species number and distribution, it is the main information necessary for the study of diversification processes (i.e., the tempo and mode of speciation and extinction),
<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity patterns are shaped across space, time and clades [@Morlon2014].
Evolutionary understanding also relies on comparative studies, for which knowing the time context for all life is crucial. Efforts to have a whole tree of life have been great and here are some examples.
<!--INTRODUCTION: 2nd paragraph-->
When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism, all of which greatly vary across different organisms and through time. <!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.

@@

Another source that has been widely used to inform about the timing of lineage origin is the relative rate of DNA or aminoacid substitution.
It is estimated from hypothesis of character homology (alignments) for reconstructing phylogenetic relationships.
<!--describe the two components of rel subst rate? bc they contain the information on absolute rate of substitution and absolute time.-->
Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--obtained in the lab?? birds-->
to <!--dissociate rel subs rates components and--> generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages. <!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin. Hence, the accumulation of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms with poor or no fossil record, but is also largely used for organisms with rich fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->
<!--INTRODUCTION: 3rd paragraph-->
In the past two decades, the possibility to obtain good quality DNA sequences
<!-- from a wide variety of organisms became a reality, which,  -->
coupled to methodological developments in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.
<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->

@@

Total ssp number estimated to be between 3 and 100 million species, as mentiones in [@Mora2011]:
ttol_spp <- 97085
(ttol_spp * 100) / 3e+6  # 3.236167
(ttol_spp * 100) / 100e+6  # 0.097085
-->
<!--
While its node-time search is very powerful, using available data to infer divergence time from lineages that have never been sequenced before, its timetree option is less powerful, their main flaw being that it does not provide source data.
However, their primary database is not fully open, giving information on a per node basis. So, it is not possible to easily access information from original studies for a group of taxa (One would need to perform a pair of taxa search for all pairs from the group of interest).
Divergence time estimates for all taxa in the TToL are precomputed and stored in the TimeTree database
All searches allow species names and higher group names. Common names are allowed in all searches but a list of taxa.
one-taxon search was not working for dog, though.
1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the age of main divergence events occuring along the lineage evolutionary history, i.e., when it shared a comon ancestor with all main lineages from which it diverged a long time ago.
It shows results only graphically on the website.
2) ttol parwise search: allows common names, species names and higher group names
    how it handles names missing from tree: it looks in the NCBI taxonomy for the closest relative of the species or higher taxa. Then uses the closest relative to get the lineage pair age from ttol.
3) ttol group of taxa
4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle search, felidae, hominidae, phrynops search, some_ants species name search. What kind of files does it accepts: txt, rtf, doc, docx?
-->

@@

Despite its great importance, analytical tools to summarize available information on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of existing data. First, original chronograms available publicly are scattered across various repositories (otol tree store, dryad, treebase, journals supplementary data) usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of data curation and knowledge on the group of interest is still necessary. For example, <!--General issues with dating techniques:--> choosing correct fossils for calibration requires a lot of expertise and knowledge on the group. An incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research community considers it as an important or even crucial step before data analysis. Hence, automated processes for large data analysis are frequently received with skepticism.
<!--INTRODUCTION: 5th paragraph-->

@@

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>

<!--(BOLD and OToL species names, are they homogeneous/standardized?-->

DateLife is a service for searching and processing information on ages of any taxon <!--or group of taxa-->of interest, available in chronograms from public data repositories of published peer reviewed studies. From these data, it can also generate new taxon age information<!--by using expert available phylogenies and scaling them to absolute time using information on ages obtained from published chronograms,--> by calling a variety of external services and tested algorithms.
It takes advatage of the rotl, ape, and geiger packages to gather, process, and present information.

It only requires a set of taxon names as input, in the form of a comma separated listing or vector, or of a phylogeny with taxon names on the tips.
Taxon names can correspond to binomial species names or clades<!--common names (will we implement it?)-->.
When taxon names are clades, DateLife pulls all accepted species names within the clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using a service of rphylotastic R package<!--citation-->.
Names belonging to subspecies or any other infraspecific category are treated as species.
DateLife can process input names with the taxon name resolution service (TNRS), <!--which "corrects misspelled names and authorities, standardizes variant spellings, and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->which corrects misspelled names or typos, and standardizes variation in spelling and synonyms [@Boyle2013], increasing the probability to correctly find the queried taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's reference taxonomy using a service from the R package rotl [@Michonneau2016]. <!--we'll implement the option to use other reference taxonomies in the future -->

DateLife's main function searches taxon names across the chronogram database specified by the user. At the moment, it queries chronograms from the OToL [@Hinchliff2015]<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repository.
DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest. It then stores taxon age information from each chronogram individually as a patristic matrix, named with the citation of the original study. This format allows a rapid summary in a number of different ways, including: 1) citations of the original studies containing the subset chronograms, 2) a list of mrca ages of subset chronograms, 3) a list of complete subset chronograms in newick or phylo format, 4) a table containing all information retrieved in html or R's data frame format, or 5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths. <!--(how to better explain the latter?? Also, I just discovered the median gives non ultrametric trees...).-->

@@

DateLife's code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time relative to number of input names <!--number of chronograms in database--> and DateLife service.
Input processing increases roughly linearly with number of input taxon names, and increases considerably if tnrs service is activated (`r figure_runtime_main`).
<!--
Speed with different number of input lineages and types of analysis:
-->
Results show that searching time increases linearly with number of input names and number of chronograms in database.

@@

<!--Potential applications shown here-->Here, we have shown that DateLife allows an easy and fast obtention of all publicly available information on taxon ages, which can be used to generate new data. <!--Potential applications not shown here-->This information can be used to account for the effect of phylogenetic signal in studies of trait evolution; to explore potential speciation and extinction dynamics of interest within a clade; to obtain a time frame of biogeographical events; for science communication and outreach, amongst others.
<!--Why you should use it-->Compared to similar platforms such as time tree of life and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge from users for any of its functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
it facilitates reproducibility of analyses;

@@

>>>>

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
fig_workflow_main <- "Fig. 1"
fig_runtime_main <- "Fig. 2"  
# make_datelife_query_plus_get_datelife_result
```

@@

<!--INTRODUCTION-->
Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
constitutes in many ways the fundamental/main knowledge necessary for evolutionary
understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information
allowing to test alternative evolutionary hypothesis in many areas of research,
from developmental to ecological questions.
Along with migration rates and patterns, it is fundamental to test biogeographical
hypothesis-->
Coupled to species number and distribution, it is the main information necessary
for the study of diversification processes (i.e., the tempo and mode of speciation
and extinction),
<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
patterns are shaped across space, time and clades [@Morlon2014].
Evolutionary understanding also relies on comparative studies, for which knowing
the time context for all life is crucial. Efforts to have a whole tree of life have
been great and here are some examples.
<!--INTRODUCTION: 2nd paragraph-->
When organisms are preserved in a fossil form, a time frame of taxon origin can
be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a
number of parameters including the nature of the habitat, population size, species
range breadth and physical characteristics of the organism, all of which greatly
vary across different organisms and through time.
<!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence
<!--taxon ages--> for all life is<!--, alas,--> not possible.

@@

Another source that has been widely used to inform about the timing of lineage origin
is the relative rate of DNA or aminoacid substitution.
It is estimated from hypothesis of character homology (alignments) for reconstructing
phylogenetic relationships.
<!--describe the two components of rel subst rate? bc they contain the information
on absolute rate of substitution and absolute time.-->
Molecular dating techniques use external data such as absolute time calibrations
(e.g., fossils, geologic events) or absolute substitution rates <!--obtained in
the lab?? birds-->to <!--dissociate rel subs rates components and--> generate dated
phylogenies (chronograms) which contain information on absolute times of node divergence
and taxon ages. <!--With molecular dating methods, an external calibration point
is still necessary to estimate the absolute time of origin. Hence, the accumulation
of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms
with poor or no fossil record, but is also largely used for organisms with rich
fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the
characteristics of the fossilization process, the quality of dates inferred directly
from it is not always considered the best.-->
<!--INTRODUCTION: 3rd paragraph-->
In the past two decades, the possibility to obtain good quality DNA sequences
<!-- from a wide variety of organisms became a reality, which,  -->
coupled to methodological developments in phylogenetic and dating inference, allowed
the application of molecular dating methods on a very large amount and diversity
of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of both fossil and molecular-based data on taxon
ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased
usage by the community<!--total evidence and revbayes (fossilized birth-death)-->,
which coupled to better sharing data practices, are greatly contributing to the
accumulation in number and type of available data on taxon ages.
<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->

@@

Total ssp number estimated to be between 3 and 100 million species, as mentiones in [@Mora2011]:
ttol_spp <- 97085
(ttol_spp * 100) / 3e+6  # 3.236167
(ttol_spp * 100) / 100e+6  # 0.097085
-->
<!--
While its node-time search is very powerful, using available data to infer divergence
time from lineages that have never been sequenced before, its timetree option is
less powerful, their main flaw being that it does not provide source data.
However, their primary database is not fully open, giving information on a per node
basis. So, it is not possible to easily access information from original studies
for a group of taxa (One would need to perform a pair of taxa search for all pairs
from the group of interest).
Divergence time estimates for all taxa in the TToL are precomputed and stored in
the TimeTree database
All searches allow species names and higher group names. Common names are allowed
in all searches but a list of taxa.
one-taxon search was not working for dog, though.
1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the
age of main divergence events occuring along the lineage evolutionary history, i.e.,
when it shared a comon ancestor with all main lineages from which it diverged a
long time ago.
It only shows results graphically on the website.
2) ttol parwise search: allows common names, species names and higher group names
    how it handles names missing from tree: it looks in the NCBI taxonomy for the
    closest relative of the species or higher taxa. Then uses the closest relative
    to get the lineage pair age from ttol.
3) ttol group of taxa
4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle
search, felidae, hominidae, phrynops search, some_ants species name search. What
kind of files does it accepts: txt, rtf, doc, docx?
-->

@@

Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration r
equires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.
<!--INTRODUCTION: 5th paragraph-->

@@

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>
```{r, echo = FALSE}
fig_workflow_main_foot <- "Figure 1. Summarized DateLife workflow. Only main actions are shown."
```
<!--(BOLD and OToL species names, are they homogeneous/standardized?-->

DateLife is a service for searching and processing information on ages of any number
of taxa <!--or group of taxa-->of interest, across chronograms available in public
data repositories coming from published peer reviewed studies. It can also generate
new taxon age information<!--by using expert available phylogenies and scaling them
to absolute time using information on ages obtained from published chronograms,--> by l
inking several external services and tested algorithms.
It takes advantage of the rotl, ape, and geiger packages to gather, process, and
present information.

It only requires a set of taxon names as input, in the form of a comma separated
listing or vector, or of a phylogeny with taxon names on the tips.
Taxon names can correspond to binomial species names or clades
<!--common names (will we implement it?)-->.
When taxon names are clades, DateLife pulls all accepted species names within the
clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using
a service of rphylotastic R package<!--citation-->.
Names belonging to subspecies or any other infraspecific category are treated as
species.
DateLife can process input names with the taxon name resolution service (TNRS),
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects misspelled names or typos, and standardizes variation in spelling
and synonyms [@Boyle2013], increasing the probability to correctly find the queried
taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's
reference taxonomy using a service from the R package rotl [@Michonneau2016].
<!--we'll implement the option to use other reference taxonomies in the future -->

DateLife's main function searches taxon names across the chronogram database specified
by the user. At the moment, it queries chronograms from OToL [@Hinchliff2015]
<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from
1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel
WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive
tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repository.
DateLife identifies chronograms having at least two taxon names, and subset them
to contain only the taxa of interest. It then stores taxon age information from
each chronogram individually as a patristic matrix, named with the citation of the
original study. This format allows a rapid summary in a number of different ways,
including:
1) citations of the original studies containing the subset chronograms,
2) a list of mrca ages of subset chronograms,
3) a list of complete subset chronograms in newick or phylo format,
4) a table containing all information retrieved in html or R's data frame format, or
5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths. <!--(how to better explain the latter?? Also, I just discovered the median gives non ultrametric trees...).-->

@@

DateLife's code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time relative to number of input names
<!--number of chronograms in database--> and DateLife service.
Input processing increases roughly linearly with number of input taxon names, and
increases considerably if tnrs service is activated (`r fig_runtime_main`).
<!--
Speed with different number of input lineages and types of analysis:
-->
Results show that searching time increases linearly with number of input names and
number of chronograms in database.

@@

<!--Potential applications shown here-->Here, we have shown that DateLife allows
an easy and fast obtention of all publicly available information on taxon ages,
which can be used to generate new data. <!--Potential applications not shown here-->
This information can be used to account for the effect of phylogenetic signal in
studies of trait evolution; to explore potential speciation and extinction dynamics
of interest within a clade; to obtain a time frame of biogeographical events; for
science communication and outreach, amongst others.
<!--Why you should use it-->Compared to similar platforms such as time tree of life
and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge from users for any of its functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
it facilitates reproducibility of analyses;

@@

```{r pressure, out.width="100%", include=FALSE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
knitr::include_graphics("fig1.pdf", dpi = 360)
```
